Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism

### 2023-10-27 paper/main.qmd
Update paper

Generating an artificial dataset required a specification of the true
measurement error; we specified a standard deviation of 0.1 on logarithmic
scale for enzyme and metabolite concentration measurements, corresponding to
approximately 10% measurement error and a standard deviation for each reaction
measurement approximately 10% of the simulated value.

@@

To investigate how the flux of GNMT between conditions [X] and [Y] changes
we compared the regulatory contributions in the posterior distribution.
Presented in [@fig-decomposition top left] is the regulatory description of
GNMT, an irreversible enzyme that is homotropically activated by its substrate,
competitively inhibited by its product and heterotropically inhibited. The
diverse regulation makes it the ideal test case to elucidate regulatory changes.

Here we found that saturation and allosteric effects were the main drivers
of regulation between the two conditions. @fig-decomposition shows the distributions
of the regulatory components as the log-ratio of each component with [X] in
the numerator and [Y] in the denominator, positive values suggest that this
component was increased in experiment [X] relative to [Y], with 0 indicating no
difference. Naturally, this presents the ideal statistical test when evaluating
the regulation, with the probability of regulation given by the relative
proportion of samples above and below the 0 point.

We demonstrate that Maud can accurately reproduce the regulatory effect between
the two conditions by comparing it to the ground truth. As the true values are
well within our probability mass of the posterior we can say that Maud is able
to accurately reproduce the regulatory behaviour of metabolic networks and also
provides a tool for statistical inference of the regulatory behaviour.

Here we have demonstrated a key feature in Maud, that is, direct inference on
any parameter that arises from the data generating process. More succinctly,
we show the benefit of error propagation that Bayesian inference provides as
researchers can directly use quantiles for hypothesis testing.

Our tool allows inference directly on the joint posterior distribution rather
than the marginal parameter. The implication is that inference will be better
than that of the marginal distributions due to correlated parameter values, if
the data generating process can be directly modelled.

>>>>

Generating an artificial dataset required a specification of the true
measurement error. For enzyme and metabolite concentration measurements
we specified a standard deviation of 0.1 on natural logarithmic scale,
corresponding to approximately 10% measurement error. For each reaction
measurement a measurement standard deviation of approximately 10% of the
simulated value.

@@

To demonstrate how Maud's output can be used to yield useful metabolic insights
we used the results of our case study to explain why the flux of the enzyme
$GNMT$ is higher in condition 1 than in condition 2. GNMT is an irreversible
enzyme that is homotropically activated by its substrate, competitively
inhibited by its product and heterotropically inhibited. The diverse regulation
makes it the ideal test case to elucidate regulatory changes.

@fig-decomposition shows the regulatory description of GNMT, according to the
results our main case study analysis. Each curve shows the marginal posterior
distribution of the ratio of the corresponding regulatory component in condition
1 compared with condition 2. A positive value indicates that the component
was increased in condition 1 relative to condition 2, with 0 indicating no
difference. The probability, according to our model, of the component acting in
each direction is given by the relative area under the curve on each side of the
0 point.

Our model correctly inferred that saturation and allosteric effects were the
main drivers of regulation between the two conditions in this case, with the
curves for each component aligning with the ground truth shown in red.

Importantly, this form of analysis takes into account all modelled sources of
uncertainty, including uncertainty about the true values of the flux in each
condition, and propagates this uncertainty into the final conclusion. Our result
shows that Maud could be used in this realistic case not only to provide a user
with an explanation for an observed difference in fluxes, but also a reasonable
judgement as to the explanation's robustness.
